---
title: "Map"
---

You can download this .qmd file from [here](https://github.com/joeroith/264_spring_2026/blob/main/12_maps.qmd).  Just hit the Download Raw File button.


```{r}
#| message: false
#| warning: false

# Initial packages required (we'll be adding more)
library(tidyverse)
library(maps)
library(lubridate)
library(leaflet)
library(sf)
library(rvest)
library(viridis)
library(mdsr) 
library(htmltools)
# package associated with our MDSR book
```


### On Your Own

1. Let's return to our example from above where we recreated an [interactive choropleth map](https://rstudio.github.io/leaflet/articles/choropleths.html) of population densities by US state.  Recall how that plot was very suspicious?  The population density data that came with the state geometries from [our source](https://rstudio.github.io/leaflet/json/us-states.geojson) seemed incorrect. 

In `09_table_scraping.qmd` (On Your Own #1) we went out and found the real statewise population densities (from [here](https://en.wikipedia.org/wiki/List_of_states_and_territories_of_the_United_States_by_population_density)) and created a tidy data frame.  Now let's merge that with our state geometry shapefiles and regenerate our interactive choropleth map.

```{r}
# Code from OYO #1 in 09_table_scraping.qmd

library(rvest)
library(polite)

session <- bow("https://en.wikipedia.org/wiki/List_of_states_and_territories_of_the_United_States_by_population_density", force = TRUE)

result <- scrape(session) |>
  html_nodes(css = "table") |> 
  html_table(header = TRUE, fill = TRUE)
density_table <- result[[1]]
density_table

density_data <- density_table |>
  select(1, 2, 4, 5) |>
  filter(!row_number() == 1) |>
  rename(Land_area = `Land area`) |>
  mutate(state_name = str_to_lower(as.character(Location)),
         Density = parse_number(Density),
         Population = parse_number(Population),
         Land_area = parse_number(Land_area)) |>
  select(-Location)

density_data
```

```{r}
# merge with the shapefile
state1 <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") |>
select(name, geometry)|>
mutate(state_name = str_to_lower(name))
density_map_sf <- state1 |>
left_join(density_data, by = "state_name")|>
filter(!(name %in% c( "Alaska", " Hawaii", "Puerto Rico")))
# Now I want to check
density_map_sf |>
st_drop_geometry() |>
count()
```

```{r}
# Making interactive map
# Create color bins
bins <- c(0, 10, 25, 50, 100, 250, 500, 1000, Inf)
pal <- colorBin("YlOrRd", domain = density_map_sf$Density, bins = bins)
# Create hover labels
labels <- density_map_sf |>
mutate(
label_text = str_c(
"<strong>", name, "</strong><br/>",
"Density: ", round(Density, 1), " people / sq mile<br/>",
"Population: ", scales::comma(Population), "<br/>",
"Land area: ", scales::comma(Land_area), " sq mi"
)
) |>
pull(label_text) |>
lapply(HTML)
leaflet(density_map_sf) |>
addTiles() |>
setView(-96, 37.8, 4) |>
addPolygons(
fillColor = ~pal(Density),
weight = 1,
color = "black",
opacity = 1,
fillOpacity = 0.8,
highlightOptions = highlightOptions(
weight = 3,
color = "#888",
bringToFront = TRUE
),
label = labels,
labelOptions = labelOptions(
direction = "auto",
textsize = "14px",
style = list("font-weight" = "normal", padding = "3px 8px")
)
) |>
addLegend(
pal = pal,
values = ~Density,
opacity = 0.8,
title = "Population density",
position = "bottomright"
)
```


2. The `states` dataset in the `poliscidata` package contains 135 variables on each of the 50 US states.  See [here](https://rdrr.io/cran/poliscidata/man/states.html) for more detail.

Your task is to create a *two* meaningful choropleth plots, one using a numeric variable and one using a categorical variable from `poliscidata::states`.  You should make *two* versions of each plot: a static plot using the `maps` package and `ggplot()`, and an interactive plot using the `sf` package and `leaflet()`.  Write a sentence or two describing what you can learn from each plot.

Here's some R code and hints to get you going:

```{r}
#| message: false
#| warning: false

# Get info to draw US states for geom_polygon (connect the lat-long points)
library(maps)
states_polygon <- as_tibble(map_data("state")) |>
  select(region, group, order, lat, long)

# See what the state (region) levels look like in states_polygon
unique(states_polygon$region)


# Get info to draw US states for geom_sf and leaflet (simple features object 
#   with multipolygon geometry column)
library(sf)
states_sf <- 
  read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") |>
  select(name, geometry)

# See what the state (name) levels look like in states_sf
unique(states_sf$name)


# Load in state-wise data for filling our choropleth maps
#   (Note that I selected my two variables of interest to simplify)
library(poliscidata)   # may have to install first
polisci_data <- as_tibble(poliscidata::states) |>
  select(state, carfatal07, cook_index3)

# See what the state (state) levels look like in polisci_data
unique(polisci_data$state)   # can't see trailing spaces but can see
                             #   lack of internal spaces
print(polisci_data)   # can see trailing spaces
```

R code hints:

- stringr functions like `str_squish` and `str_to_lower` and `str_replace_all` (be sure to carefully look at your keys!)
- *_join functions (make sure they preserve classes)
- filter so that you only have 48 contiguous states (and maybe DC)
- for help with colors: https://rstudio.github.io/leaflet/reference/colorNumeric.html
- be sure labels pop up when scrolling with leaflet


```{r}
# Make sure all keys have the same format before joining:
#   all lower case, no internal or external spaces
polisci_data1 <- polisci_data |>
mutate(
state_key = state |>
str_squish() |>
str_to_lower() |>
str_replace_all(" "
,
"")
)

states_polygon1 <- states_polygon |>
mutate(
state_key = region |>
str_squish() |>
str_to_lower() |>
str_replace_all(" "
,
"")
)
states_sf1 <- states_sf |>
mutate(
state_key = name |>
str_squish() |>
str_to_lower() |>
str_replace_all(" "
,
"")
)
```

```{r}
# Now we can merge data sets together for the static and the interactive plots


# Merge with states_polygon (static)
states_static <- states_polygon1 |>
left_join(polisci_data1, by = "state_key") |>
filter(!(region %in% c( "Alaska", " Hawaii", "Puerto Rico")))
# Check that merge worked for 48 contiguous states
states_static |>
distinct(region)|>
count()
# Check that merge worked for 48 contiguous states

# Merge with states_sf (static or interactive)
states_interactive <-states_sf1 |>
left_join(polisci_data1, by = "state_key") |>
filter(!(name %in% c("Alaska", " Hawaii", " Puerto Rico")))
# Check that merge worked for 48 contiguous states
states_interactive |>
st_drop_geometry()|>
count()
```


Numeric variable (static plot):

```{r}
ggplot(states_static, aes(x = long, y = lat, group = group)) +
geom_polygon(aes(fill = carfatal07), color = "white", linewidth = 0.2) +
coord_map() +
theme_void() +
scale_fill_viridis_c(na.value = "grey90") +
labs(fill = "Car fatalities\n(2007)",
title = "Motor vehicle fatality rates by state")
```


Numeric variable (interactive plot):

```{r}
pal_num <- colorNumeric(
palette = "viridis",
domain = states_interactive$carfatal07,
na.color = "grey90"
)
labels_num <- states_interactive |>
mutate(
label_text = str_c(
"<strong>", name, "</strong><br/>",
"Car fatalities (2007): ", round(carfatal07, 2)
)
) |>
pull(label_text) |>
lapply(HTML)
leaflet(states_interactive) |>
addTiles() |>
setView(-96, 37.8, 4) |>
addPolygons(
fillColor = ~pal_num(carfatal07),
weight = 1,
color = "white",
opacity = 1,
fillOpacity = 0.8,
label = labels_num,
highlightOptions = highlightOptions(
weight = 3,
color = "#666",
bringToFront = TRUE
)
)
```


Categorical variable (static plot):

```{r}
# be really careful with matching color order to factor level order
states_interactive |>
st_drop_geometry() |>
count(cook_index3)

# Make factor levels explicit so color order stays consistent
states_static <- states_static |>
mutate(cook_index3 = factor(cook_index3))
ggplot(states_static, aes(x = long, y = lat, group = group)) +
geom_polygon(aes(fill = cook_index3), color = "black", linewidth = 0.2) +
coord_map() +
theme_void() +
scale_fill_brewer(palette = "RdBu", na.value = "grey90") +
labs(fill = "Cook index",
title = "State partisan leaning")
```


Categorical variable (interactive plot):

```{r}
# may use colorFactor() here
# may use colorFactor() here
states_interactive <- states_interactive |>
mutate(cook_index3 = factor(cook_index3))
pal_cat <- colorFactor(
palette = "RdBu",
domain = states_interactive$cook_index3,
na.color = "grey90"
)
labels_cat <- states_interactive |>
mutate(
label_text = str_c(
"<strong>", name, "</strong><br/>",
"Cook index: ", cook_index3
)
) |>
pull(label_text) |>
lapply(HTML)
leaflet(states_interactive) |>
addTiles() |>
setView(-96, 37.8, 4) |>
addPolygons(
fillColor = ~pal_cat(cook_index3),
weight = 1,
color = "white",
opacity = 1,
fillOpacity = 0.8,
label = labels_cat,
highlightOptions = highlightOptions(
weight = 3,
color = "#111",
bringToFront = TRUE
)
) |>
addLegend(
pal = pal_cat,
values = ~cook_index3,
title = "Cook index",
position = "bottomright"
)
```


