---
title: "Map"
editor_options: 
  chunk_output_type: console
---

```{r, echo = FALSE}
#| message: false
#| warning: false
#| results: 'hide'

# Initial packages required (we'll be adding more)
library(tidyverse)
library(maps)
library(lubridate)
library(leaflet)
library(sf)
library(rvest)
library(viridis)
library(mdsr) 
library(htmltools)
# package associated with our MDSR book
```

The `states` dataset in the `poliscidata` package contains 135 variables on each of the 50 US states. See [here](https://rdrr.io/cran/poliscidata/man/states.html) for more detail.

In this tab, I will create a *two* meaningful choropleth plots, one using a numeric variable and one using a categorical variable from `poliscidata::states`. You will see *two* versions of each plot: a static plot using the `maps` package and `ggplot()`, and an interactive plot using the `sf` package and `leaflet()`. Finally, I will write a sentence or two describing what I learn from each plot.

```{r, echo=FALSE}
#| message: false
#| warning: false
#| results: 'hide'
# Get info to draw US states for geom_polygon (connect the lat-long points)
library(maps)
states_polygon <- as_tibble(map_data("state")) |>
  select(region, group, order, lat, long)

# See what the state (region) levels look like in states_polygon
unique(states_polygon$region)


# Get info to draw US states for geom_sf and leaflet (simple features object 
#   with multipolygon geometry column)
library(sf)
states_sf <- 
  read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") |>
  select(name, geometry)

# See what the state (name) levels look like in states_sf
unique(states_sf$name)


# Load in state-wise data for filling our choropleth maps
#   (Note that I selected my two variables of interest to simplify)
library(poliscidata)   # may have to install first
polisci_data <- as_tibble(poliscidata::states) |>
  select(state, carfatal07, cook_index3)

# See what the state (state) levels look like in polisci_data
unique(polisci_data$state)   # can't see trailing spaces but can see
                             #   lack of internal spaces
print(polisci_data)   # can see trailing spaces
```


```{r, echo=FALSE}
#| results: 'hide'
# Make sure all keys have the same format before joining:
#   all lower case, no internal or external spaces
polisci_data1 <- polisci_data |>
mutate(
state_key = state |>
str_squish() |>
str_to_lower() |>
str_replace_all(" "
,
"")
)

states_polygon1 <- states_polygon |>
mutate(
state_key = region |>
str_squish() |>
str_to_lower() |>
str_replace_all(" "
,
"")
)
states_sf1 <- states_sf |>
mutate(
state_key = name |>
str_squish() |>
str_to_lower() |>
str_replace_all(" "
,
"")
)
```

```{r, echo=FALSE}
#| results: 'hide'
# Now we can merge data sets together for the static and the interactive plots


# Merge with states_polygon (static)
states_static <- states_polygon1 |>
left_join(polisci_data1, by = "state_key") |>
filter(!(region %in% c( "Alaska", " Hawaii", "Puerto Rico")))
# Check that merge worked for 48 contiguous states
states_static |>
distinct(region)|>
count()
# Check that merge worked for 48 contiguous states

# Merge with states_sf (static or interactive)
states_interactive <-states_sf1 |>
left_join(polisci_data1, by = "state_key") |>
filter(!(name %in% c("Alaska", " Hawaii", " Puerto Rico")))
# Check that merge worked for 48 contiguous states
states_interactive |>
st_drop_geometry()|>
count()
```

### Numeric variable (static plot):

```{r, echo=FALSE}
ggplot(states_static, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = carfatal07), color = "white", linewidth = 0.2) +
  coord_map() +
  theme_void() +
  scale_fill_viridis_c(
    option = "viridis",
    na.value = "grey90",
    name = "Fatalities\nper 100k"
  ) +
  labs(
    title = "Motor Vehicle Fatality Rates by U.S. State (2007)",
    caption = "Source: https://rdrr.io/cran/poliscidata/man/states.html"
  )
```

*Motor vehicle fatality rates vary significantly across states, with higher rates concentrated in the South and some Mountain West states. In contrast, states in the Northeast and on the West Coast tend to have lower fatality rates. This pattern may reflect differences in population density, road conditions, and driving behavior, as rural areas often experience higher speeds and longer emergency response times.*

### Numeric variable (interactive plot):

```{r, echo=FALSE}
pal_num <- colorNumeric(
palette = "viridis",
domain = states_interactive$carfatal07,
na.color = "grey90"
)
labels_num <- states_interactive |>
mutate(
label_text = str_c(
"<strong>", name, "</strong><br/>",
"Car fatalities (2007): ", round(carfatal07, 2)
)
) |>
pull(label_text) |>
lapply(HTML)
leaflet(states_interactive) |>
addTiles() |>
setView(-96, 37.8, 4) |>
addPolygons(
fillColor = ~pal_num(carfatal07),
weight = 1,
color = "white",
opacity = 1,
fillOpacity = 0.8,
label = labels_num,
highlightOptions = highlightOptions(
weight = 3,
color = "#666",
bringToFront = TRUE
)
) |>
  addLegend(
    pal = pal_num,
    values = ~carfatal07,
    title = "Fatalities per 100,000",
    position = "bottomright"
  )
```

*The interactive feature allows users to hover over each state to view precise fatality rates, providing more detailed insight than the static map.*

### Categorical variable (static plot)

```{r}
#| echo: false
# be really careful with matching color order to factor level order

# Make factor levels explicit so color order stays consistent
states_static <- states_static |>
mutate(cook_index3 = factor(cook_index3))
ggplot(states_static, aes(x = long, y = lat, group = group)) +
geom_polygon(aes(fill = cook_index3), color = "black", linewidth = 0.2) +
coord_map() +
theme_void() +
scale_fill_brewer(palette = "RdBu", na.value = "grey90") +
  labs(
    title = "State Partisan Leaning by U.S. State",
    caption = "Source: https://rdrr.io/cran/poliscidata/man/states.html"
  )
```

*This map shows which political party each state tends to support. The pattern shows that many coastal states lean Democratic, while many states in the South, Midwest, and Mountain West lean Republican.*

### Categorical variable (interactive plot):

```{r, echo=FALSE}
# may use colorFactor() here
# may use colorFactor() here
states_interactive <- states_interactive |>
mutate(cook_index3 = factor(cook_index3))
pal_cat <- colorFactor(
palette = "RdBu",
domain = states_interactive$cook_index3,
na.color = "grey90"
)
labels_cat <- states_interactive |>
mutate(
label_text = str_c(
"<strong>", name, "</strong><br/>",
"Cook index: ", cook_index3
)
) |>
pull(label_text) |>
lapply(HTML)
leaflet(states_interactive) |>
addTiles() |>
setView(-96, 37.8, 4) |>
addPolygons(
fillColor = ~pal_cat(cook_index3),
weight = 1,
color = "white",
opacity = 1,
fillOpacity = 0.8,
label = labels_cat,
highlightOptions = highlightOptions(
weight = 3,
color = "#111",
bringToFront = TRUE
)
) |>
addLegend(
pal = pal_cat,
values = ~cook_index3,
title = "Cook index",
position = "bottomright"
)
```

*By hovering over the states, users can see the exact category for each state, making it easier to compare regional political patterns across the country.*
